Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-22 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
moving the large corner plot, added text related to the small one. Working on appendices

Let us also contrasts our results with those of Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}
which relied on the BBH inspiral to constrain $\ell_{\rm EdGB}$, as discussed in Sec.~\ref{sec:review_edgb}.
%
We see that EdGB gravity provides an example of a theory in which, with current GW events, the inspiral portion
of the signal can be more constraining than the ringdown portion of the signal.
%
Two reasons together can explain our negative results. First, as observed by
Ref.~\cite{Blazquez-Salcedo:2016enn}, the QNMs of EdGB BHs only differ slightly
from their Schwarzschild counterparts. Second, the larger mass $M_{\rm f}$ of the
remnant BH, suppresses scalar field's charge relatively to the initial binary
components.
%
As we will next, the situation is the opposite for dCS gravity.

@@

\hscomm{Alessandra, I moved the plot here as we discussed.}
%
In Fig.~\ref{fig:corner_plot}, we show the final spin $\chi_{\rm f}$ and
remnant mass $M_{\rm f}$ for GW150914 for GR and dCS gravity.
%
We see that indeed our waveform models does not introduce substantial changes
to the GR estimates on these parameters, as required by the ParSpec expansion
(see discussion in Sec.~\ref{sec:review_parspec}).
%
We observed the same in all other theories we studied.

@@

Let us now consider the quartic EFT of GR, as our final example.
%
In Fig.~\ref{fig:qEFT_exec_sum} we show the posteriors on $\ell_{\rm qEFT}$ for GW150914 (top panel),
GW200129 (middle panel) for $N_{\rm max} = 0$, which are qualitatively similar to the cubic EFT of GR.
%
We find that both the EFT~\eqref{eq:eft_bound} and ParSpec~\eqref{eq:parspec_bound} are satisfied,
hence allowing us to infer bounds on $\ell_{\rm qEFT}$.
%
As Fig.~\ref{fig:qEFT_cdf} shows the CDFs for both events reach 90\% for $\gm = M_{\rm f}$ before $\varepsilon = 1/2$.
%
Consequently, from the joint posterior of $\ell_{\rm qEFT}$ we quote our final result,
%
\begin{equation}
    \ell_{\rm qEFT} \leqslant 51.3~\textrm{km} \quad \textrm{at 90\% credible level}\,.
\end{equation}
%
% \abcomm{Please update the paragraph below.}
In this theory, we have considered only $N_{\rm max} = 0$.
%
\hs{We found that the inclusion of the spin corrections produced waveforms which for $\ell_{\rm qEFT} \gtrsim 80$~km
had a ringdown segment larger (sometimes seconds long) that the inspiral-plugne segment.}
% for which the ringdown portions of the larger than the inspiral-plunge part. \abcomm{The last sentence
% has some problems in English. I do not udnerstand it.}
%
To avoid this issue we have lowered the value of $\ell^{\rm max}_{\rm qEFT}$, but by doing so we have obtained posteriors
which were essentially uniform, as our prior, and thus uninformative. Hence, we do not quote any results for $N_{\rm max} = 1$.
%
Table~\ref{tab:summary_qeft} summarizes our findings for the quartic EFT of GR.
%
\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/qeft_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function for the coupling constant $\ell_{\rm qEFT}$ for
GW150914 (top panel), GW200129 (middle panel) and combined events (bottom panel).
}
\label{fig:qEFT_exec_sum}
\end{figure}

@@

We found, in particular, that within the interpretation these theories as EFTs and the
region of validity of the ParSpec framework, the coupling constant of dCS gravity is bound to
$\ell_{\rm dCS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129.
%
In contrast, we could not place any constraint of the coupling constants in EdGB gravity $\ell_{\rm EdGB}$.
%
This dichotomy between the two theories has an analogy with works that considered
the inspiral part of the GW signal alone.
%
In that case, it is dCS gravity which cannot be constrained with current GW
events, while EdGB can, as shown in Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
% HS: changed my mind -- I think it would divert from the original flow.
% \hscomm{I will add a sentence here on Refs.~\cite{Alexander:2017jmt,Loutrel:2018ydv,Loutrel:2018rxs}.}
%
We hope that our results serve as further motivation to address these
challenging problems.

@@

Our work shows that there are theories of gravity (such as dCS) can be unconstrained by studying the
inspiral portion of GW events, {\it yet} can be constrained, if we analyze the ringdown.
%
Our results show that to fulfill the potential to test GR with GW observations,
further work on modeling of the late-inspiral-(plunge)-merger and ringdown of
coalescing BHs  (with NR simulations) and, complementary, the calculations of
QNMs of rapidly-rotating BHs in modified gravity theories (with BH perturbation
theory) are necessary.

@@

>>>>

Let us also contrasts our results with those of Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}
which relied on the BBH inspiral to constrain $\ell_{\rm EdGB}$, as discussed in Sec.~\ref{sec:review_edgb}.
%
We see that EdGB gravity provides an example of a theory in which, with current GW events, the inspiral portion
of the signal can be more constraining than the ringdown portion of the signal.
%
Two reasons together can explain our negative results. First, as observed by
Ref.~\cite{Blazquez-Salcedo:2016enn}, the QNMs of EdGB BHs only differ slightly
from their Schwarzschild counterparts. Second, the larger mass $M_{\rm f}$ of the
remnant BH, suppresses scalar field's charge relatively to the initial binary
components.
%
% As we will next, the situation is the opposite for dCS gravity.

@@

\hscomm{Alessandra, I moved the plot here as we discussed.}
%
As an additional check, to verify the robustness of our constraint, we show in
Fig.~\ref{fig:corner_plot}, the final spin $\chi_{\rm f}$ and remnant mass
$M_{\rm f}$ for GW150914 for GR and dCS gravity.
%
We see that our \pSEOB waveform model does not introduce substantial changes
to the GR estimates on these parameters, as required by the ParSpec expansion
(see discussion in Sec.~\ref{sec:review_parspec}).
%
In fact, we observed no bias on the estimation of $\Mf$ and $\chi_{\rm f}$
for all theories considered here.
%
In Appendix~\ref{app:big_contour} we show how all other intrinsic parameter parameters
remain unbiased.

@@

Let us now consider the quartic EFT of GR, as our final example.
%
In Fig.~\ref{fig:qEFT_exec_sum} we show the posteriors on $\ell_{\rm qEFT}$ for GW150914 (top panel),
GW200129 (middle panel) for $N_{\rm max} = 0$, which are qualitatively similar to the cubic EFT of GR.
%
We find that both the EFT~\eqref{eq:eft_bound} and ParSpec~\eqref{eq:parspec_bound} are satisfied,
hence allowing us to infer bounds on $\ell_{\rm qEFT}$.
%
As Fig.~\ref{fig:qEFT_cdf} shows the CDFs for both events reach 90\% for $\gm = M_{\rm f}$ before $\varepsilon = 1/2$.
%
Consequently, from the joint posterior of $\ell_{\rm qEFT}$ we quote our final result,
%
\begin{equation}
    \ell_{\rm qEFT} \leqslant 51.3~\textrm{km} \quad \textrm{at 90\% credible level}\,.
\end{equation}
%
% \abcomm{Please update the paragraph below.}
In this theory, we have considered only $N_{\rm max} = 0$.
%
\hs{We found that the inclusion of the spin corrections produced waveforms which for $\ell_{\rm qEFT} \gtrsim 80$~km
had a ringdown segment larger (sometimes seconds long) that the inspiral-plunge segment in the detectors frequency bands.}
% for which the ringdown portions of the larger than the inspiral-plunge part. \abcomm{The last sentence
% has some problems in English. I do not udnerstand it.}
%
To avoid this issue we have lowered the value of $\ell^{\rm max}_{\rm qEFT}$, but by doing so we have obtained posteriors
which were flat, just as our prior, and were thus uninformative. Hence, we do not quote any results for $N_{\rm max} = 1$.
%
Table~\ref{tab:summary_qeft} summarizes our findings for the quartic EFT of GR.
%
\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/qeft_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function for the coupling constant $\ell_{\rm qEFT}$ for
GW150914 (top panel), GW200129 (middle panel) and combined events (bottom panel).
}
\label{fig:qEFT_exec_sum}
\end{figure}

@@

We found, in particular, that within the interpretation these theories as EFTs and the
region of validity of the ParSpec framework, the coupling constant of dCS gravity is bound to
$\ell_{\rm dCS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129.
%
In contrast, we could not place any constraint of the coupling constants in
EdGB gravity $\ell_{\rm EdGB}$.
%
This dichotomy between the two theories has an analogy with works that considered
the inspiral part of the GW signal alone.
%
In that case, it is dCS gravity which cannot be constrained with current GW
events, while EdGB can, as shown in Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
This result also shows how the inclusion of theory-specific information into the ParSpec framework
can result in different outcomes for different theories, even if they predict same exponent $p$.

% HS: changed my mind -- I think it would divert from the original flow.
% \hscomm{I will add a sentence here on Refs.~\cite{Alexander:2017jmt,Loutrel:2018ydv,Loutrel:2018rxs}.}

@@

Our work shows that there are theories of gravity (such as dCS) can be unconstrained by studying the
inspiral portion of GW events, {\it yet} can be constrained, if we analyze the ringdown.
%
Our results show that to fulfill the potential to test GR with GW observations,
further work on modeling of the late-inspiral-(plunge)-merger and ringdown of
coalescing BHs  (with NR simulations) and, complementary, the calculations of
QNMs of rapidly-rotating BHs in modified gravity theories (with BH perturbation
theory) are necessary.
%
We hope that our results serve as further motivation to address these
challenging, and outstanding, problems.

@@

\section{Intrinsic binary parameters}
\label{app:big_contour}

In this Appendix, we show in Fig.~\ref{fig:corner_plot_all} a corner plot with all intrinsic binary parameters from our parameter estimation for GW150914
using the \pSEOB{} waveform models for GR and dCS.
%
We see that median of the distribution is not affected substantially by the inclusion of the non-GR parameters.
%
That is the case for $\Mf$ and $\chi_{\rm f}$ is an underlying assumption of the ParSpec framework.
%
We qualitative similar results for GW200129 and the other modified gravity theories considered in our work.
